Understanding Recovery from Malnutrition in Early Life:\\ A Multi-Omic Analysis of the Infant Gut Microbiome, Plasma Lipidome, and Neurodevelopmental Outcomes

### 2025-08-18 writing/manuscript/main.tex
Progress for the mefisto run

\newpage
\section{Abstract}
Recovery from undernutrition during childhood has historically centered on short term weight gain with little attention given to redressing the delays in neurocognitive development.
Herein we discuss the results of a prospective, randomized controlled trial for the use of an enhanced, brain-targeted micronutrient supplemented nutritional rehabilitation regimen versus the standard of care and healthy age-matched controls on measures of cognition and emotional regulation in a cohort of one-year-old Bangladeshi children.
This longitudinal, multimodal study explains biological mechanism through network analysis to understand the impacts of a dietary intervention.

\newpage
\section{Main}
% Definition of MAM
Malnutrition refers to deficiencies, excesses, or imbalances in a person’s intake of energy and/or nutrients and can refer to undernutrition, overweight, or obesity.
% Importance of MAM
Malnutrition is a global crisis (UN SDG2: Zero Hunger) \cite{UN_SDG2}.
% Why MAM and not SAM
Moderate Acute Malnutrition (MAM) is the most common, most treatable form of undernutrition.
% What does MAM do
MAM in childhood delays the development of multiple systems in the body from the gut microbiome to neurocognition.
% Treament rates
A small proportion of children with MAM are treated.
% What is treatment
Treatment involves refeeding.
% Problems with refeeding
Current Refeeding methods focus solely on short term weight gain, associate with serious adverse events (such as diarrhoea or Refeeding Syndrome), and are often ineffective at restoring developmental trajectories.
Those that recover their weight often have lasting neurological deficits including cognition and emotional regulation.
% Microbiome connection
MAM is associated with gut microbiome immaturity.
% Microbiome targetted refeeding
Novel prototypical refeeding methods, such as microbiota-directed complementary foods (MDCFs) target the gut-microbiome-brain axis as a mediator to address these deficits.
% Effectiveness of MCDF
By inoculating gnotobiotic pig and mice preclinical models with age-characteristic gut microorganisms, MDCFs increased the levels of biomarkers and mediators of growth, bone formation, neurodevelopment, and immune function toward a state resembling healthy children \cite{gehrig2019effects}.
% Need to improve on these methods
Developing these methods into sustainable refeeding strategies is needed urgently.
% The difficulty in predicting Recovery measurements
% Our study aims
Herein we discuss the results of a study that empoys The Enhanced Ready to Use Supplementary Food (E-RUTF) plus Small Quantity Lipid Based Nutrient Supplement (E-SQLNS) as compared to standard RUTF commonly used to treat childhood Moderate Acute Malnutrition (MAM)

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=1.2]{figures/anthrochange_simple.pdf}
	\caption[Measures of cognition were not significantly impacted by refeeding]{
		Measures of cognition were not significantly impacted by refeeding.
		a) Lineplot of how children change in WLZ/WHZ during refeeding.
		b) Boxplot of days to recovery compared between the feeding groups.
	}
	\label{Figure1}
\end{figure}

\subsection{What influenced recovery}
% A background to the data collected
Acknowledging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
% risk factors for sustained recovery
% are the same factors important for reocvery rates/days to recovery/sustained recovery?
% Why mixed models - maaslin3

>>>>

\section{Abstract}
Recovery from undernutrition during childhood has historically centered on short term weight gain with little attention given to redressing the delays in neurocognitive development.
Herein we discuss the results of a prospective, randomized controlled trial for the use of an enhanced, brain-targeted micronutrient supplemented nutritional rehabilitation regimen versus the standard of care and healthy age-matched controls on measures of cognition and emotional regulation in a cohort of one-year-old Bangladeshi children.
This longitudinal, multimodal study explains biological mechanism through network analysis to understand the impacts of a dietary intervention.

\section{Main}
% Definition of MAM
Malnutrition refers to deficiencies, excesses, or imbalances in a person’s intake of energy and/or nutrients and can refer to undernutrition, overweight, or obesity.
% Importance of MAM
Malnutrition is a global crisis (UN SDG2: Zero Hunger) \cite{UN_SDG2}.
% Why MAM and not SAM
Moderate Acute Malnutrition (MAM) is the most common, most treatable form of undernutrition.
% What does MAM do
MAM in childhood delays the development of multiple systems in the body from the gut microbiome to neurocognition.
% Treament rates
A small proportion of children with MAM are treated.
% What is treatment
Treatment involves refeeding.
% Problems with refeeding
Current Refeeding methods focus solely on short term weight gain, associate with serious adverse events (such as diarrhoea or Refeeding Syndrome), and are often ineffective at restoring developmental trajectories.
Those that recover their weight often have lasting neurological deficits including cognition and emotional regulation.
% Microbiome connection
MAM is associated with gut microbiome immaturity.
% Microbiome targetted refeeding
Novel prototypical refeeding methods, such as microbiota-directed complementary foods (MDCFs) target the gut-microbiome-brain axis as a mediator to address these deficits.
% Effectiveness of MCDF
By inoculating gnotobiotic pig and mice preclinical models with age-characteristic gut microorganisms, MDCFs increased the levels of biomarkers and mediators of growth, bone formation, neurodevelopment, and immune function toward a state resembling healthy children \cite{gehrig2019effects}.
% Need to improve on these methods
Developing these methods into sustainable refeeding strategies is needed urgently.
% The difficulty in predicting Recovery measurements
% Our study aims
Herein we discuss the results of a study that empoys The Enhanced Ready to Use Supplementary Food (E-RUTF) plus Small Quantity Lipid Based Nutrient Supplement (E-SQLNS) as compared to standard RUTF commonly used to treat childhood Moderate Acute Malnutrition (MAM)

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=1.2]{figures/recovery2.pdf}
	\caption[Study design and changes in weight-for-length z-scores (WLZ) in children with moderate acute malnutrition (MAM) receiving enhanced or local refeeding]{
		Study design and changes in weight-for-length z-scores (WLZ) in children with moderate acute malnutrition (MAM) receiving enhanced or local refeeding.
		a) study design for the evaluation of children with MAM between 12 and 24 months of age, randomized to either an enhanced refeed and supplement arm (ERUSF followed by ESQLNS) or a local refeed and supplement arm (RUSF followed by SQLNS), alongside a well-nourished control group. Baseline (T1) and post-refeeding (T2) sampling included anthropometry, stool microbiome, and plasma metabolites, with additional one time measures of household, educational, and genetic covariates. Recovery was defined as WLZ \textless{} –1 SD.
		b) individual trajectories of WLZ over 52 weeks, with rapid catch-up growth during the first 15 weeks and stratification by recovery status.
		c) proportion of participants achieving recovery, sustaining recovery at 24 months, or showing no recovery in each MAM treatment arm.
		d) distribution of time to recovery between treatment arms, with a significant difference (\textit{q} \textless{} 0.05) between local RUSF and ERUSF groups. Box plots show the median, the 25th - 75th percentiles (IQR), and whiskers extending to 1.5 x IQR.
	}
	\label{Figure1}
\end{figure}

\subsection{What influenced recovery}
% A background to the data collected
Acknowledging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
% Prediction of recovery from baseline facors - MOFA
% risk factors for sustained recovery
% are the same factors important for reocvery rates/days to recovery/sustained recovery?
% Why mixed models - maaslin3
